\newpage

\begin{center}
    {\large \bf Acknowledgements }
\end{center}

\vspace{-0.2cm}


The work presented here is the tip of an iceberg, constructed with the
guidance, compassion, and effort of many. After six years in New York City,
building towards this body of work, I have a lot of people to thank; so
I will first thank you for forgiving my lack of brevity.


First, and foremost are my advisors and mentors Zolt\'an Haiman and Janna
Levin. I hold a deep respect for you both. I hope our collaborations continue
for a long time. Thank you for listening to me, pushing me, and for giving up
any sleep you might have lost for my sake.

% (also I need you to keep correcting my grammar Janna)


I have been lucky to have had a number of influential mentors whom I must thank for
getting me to where I am today. Without my high school teachers Thomas Gill,
and Brent Criswell, I don't think I would have ever realized in time that
physics is something that I love, and that I had the option of pursuing. In
college, Susan Esch gave me the confidence and positive influence that I
needed to succeed. To her and the math and physics departments at Juniata
College, namely James White, Mark Pearson, Norman Siems,
Barry Bruce, James Borgardt, and Mary Atchley, I owe a great deal. I must also 
thank those who run the NSF-REU program at the Harvard -Smithsonian center for astrophysics, 
and especially my advisor there, Greg Dobler, for the great experience they afforded me during my
first astrophysics research project.

%Caleb Wheeler

After college I had the privilege to work with Prasenjit Saha at the
University of Z\"urich, someone who taught me patience (though one can always
learn more) and what pure thought can do. I remember a trip Prasenjit and
I took to Konstanz, Germany where a long walk helped me to decide to focus on
the research that is now the subject of this thesis. Many others influenced my time
in Z\"urich, and though I cannot fit all of their names, I would like to
thank, most importantly, Raymond Angelil and Theodosia Charpidou, 
%- I've never enjoyed myself doing physics more than with Ray and I have not had many better parties than with he and Thea. 
Doug Potter, Jonathan Coles,
Antoine Klein, Cedric, Simone Callegari, Davide Martizzi, Marina
Galvani, Ben Moore, Lucio Mayer, Roman Teyssier, Joachim
Stadel, George Lake, Miguel Pato, Christine Moran, and Tina Wentz.

At Columbia I was influenced by many, my fellow graduate students with
which long discussions and many happy hours I endured, you know who you are, 
except for David Hendel (and he knows what I mean). 
%Adrian Price-Whelan,
%Jeff Andrews, Lia Corrales, Destry Saul, Cameron Hummels, Duane Lee, Erika
%Hamden, Josh Tan, Taka Tanaka, Jana Grcevich, Niel Zimmerman, Andrea
%Derdzinski, Sarah Pearson, Munier Salem, Susan Clark, Aleksay Generozov,
%Steven Mohammed, Davide Jaimes, Ron Tso, and never David Hendel (he knows what
%I mean). 
Thank you to my colleagues Brian Farris and Paul Duffel who taught me a great
deal about numerical hydrodynamics and to Caleb Wheeler, Geoffrey Ryan, Matt Turk, Jeff
Oishi, Nick Stone, Larry Price, Adrian Price-Whelan, Jeff Andrews, and Andreas
K\"upper for many interesting discussions. Thanks to Summer Ash, Marcel Ag\"ueros,
and Dave Deutsch for the the only times I actually used a telescope in
graduate school, those great evenings with the Columbia outreach and Rooftop Variables
programs. Thanks to all of the Columbia Astronomy faculty, as well as Andrew
MacFadyen, Lam Hui, Andrei Beloborodov, Brian Metzger, and Kristen Menou, all
of whom mentored me along the way. Thank you to Millie, Ayoune, and Tom
Tarduogno without whom I would have never made it through the loans,
paperwork, and bureaucracy.

I could not have done without my friends these past 6 years. Thanks for the
late nights and long trips John O'Donnell, Matt Wilde, Ben Gilmour, Adam
Bisogne, Marina Harnik, Andrea Derdzinski, Adrian Price-Whelan, Jeff Andrews, Josh Bass,
Duane Lee, Sarah Pearson, David Muirhead, Lia Corrales, Kevan Crotan, and many more...

Finally, I thank my family who has been behind me every moment. I thank my
parents John and Pamela D'Orazio for encouraging me to be curious, for
teaching me to be open minded, and for always supporting me wherever that
curiosity has lead. I thank my sister Nickie for getting me that apartment in
Harlem where we certainly had some times, and for her ability to speak my
language, to whatever end. Thanks to my grandparents John J. and Barbara
D'Orazio, and Ray and Peggy Fryling, I could not have done without their
support. Thanks to my Aunt Pat Payne RIP, Aunt Patty and Uncle Steve, and my
best friends and brothers Alex and Ryan Patterson - may the Growler never
sink.













% other
% Things:
% DD lounge
% The woods 
% The desert
% The Growler



% Finally thank you to the NSF for granting me a NSF graduate reserach fellowship for which supported me for half of my time at Columbia.




\vspace{1.0cm}
May 2016, New York, NY

%\setlength{\baselineskip}{1.11111 \baselineskip}
